---
title: "Airline Route Profitability"
output: html_document
date: "2026-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The dataset used in this project is the "Airline Route Profitability Dataset"
from Kaggle, available at: https://www.kaggle.com/datasets/waleedfaheem/airline-route-profitability-and-cost-analysis?resource=download
Author: Muhammad Waleed

```{r}
library(tidyverse)
library(tidyr)
library(lubridate)
library(patchwork)
```

```{r}
#Loading Data
airline_data <- read_csv("airline_route_profitability.csv")
glimpse(airline_data)
```

```{r}
head(airline_data)
```

```{r}
# Check missing values before
colSums(is.na(airline_data))

# Replace NAs in numeric columns with the median
airline_clean <- airline_data %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Replace NAs in text columns with "Unknown"
airline_clean <- airline_clean %>%
  mutate(across(where(is.character),
                ~ ifelse(is.na(.), "Unknown", .)))

# Check missing values after
colSums(is.na(airline_clean))

```

```{r}
glimpse(airline_clean)
```

**Key Variables:**

- **Profit**: Total revenue minus total costs for a given flight (in USD).
- **Profit_Margin**: Profit as a percentage of Total Revenue (Profit / Total_Revenue x 100). This is a fairer efficiency metric than raw Profit, since long-haul flights earn more absolute profit but may be less efficient per dollar of revenue.
- **Load_Factor**: Proportion of seats filled (Passengers / Aircraft Capacity). A value of 1.0 means the flight was completely full.
- **Route_Category**: Flight distance tier. Short Haul (1 to 3.5 hrs), Medium Haul (2.5 to 7 hrs), Long Haul (6.5 to 16.5 hrs).
- **Season**: Travel demand period. Peak (June to Aug, late Dec), Shoulder (Apr to May, Sep to Oct), Normal (Mar, Nov), Low (Jan to Feb).
- **Demand_Level**: Demand classification for a given flight (High, Medium, Low).

Note: The dataset already includes a Profit_Margin column (Profit / Total_Revenue x 100). This is a fairer efficiency metric than raw Profit, since long-haul flights earn more absolute profit but may be less efficient per dollar of revenue.

## Exploratory Analysis

Before getting into the research questions, here's a quick look at average profit across the main grouping variables.

```{r}
#Average profit by season
airline_clean %>%
  group_by(Season) %>%
  summarize(avg_profit = mean(Profit, na.rm = TRUE))
```

Profit is highest in Peak season and lowest in Low season, which makes sense since more demand generally means fuller flights and higher ticket prices.

```{r}
# Average profit and profit margin by route category
airline_clean %>%
  group_by(Route_Category) %>%
  summarize(
    avg_profit = mean(Profit, na.rm = TRUE),
    avg_margin = mean(Profit_Margin, na.rm = TRUE)
  )

#Short Haul = 1 to 3.5 hours
#Medium Haul = 2.5 to 7 hours
#Long Haul = 6.5 to 16.5 hours
```

Long-haul routes earn the most absolute profit, but the margin column tells a more nuanced story, which we explore further in the research questions.

```{r}
#Average profit and profit margin by demand level
airline_clean %>%
  group_by(Demand_Level) %>%
  summarize(
    avg_profit = mean(Profit, na.rm = TRUE),
    avg_margin = mean(Profit_Margin, na.rm = TRUE)
  )
```

High demand flights are consistently more profitable in both metrics. This is one of the clearest patterns in the dataset.

```{r}
# Convert Flight_Date to a proper date
airline_clean <- airline_clean %>%
  mutate(Flight_Date = ymd(Flight_Date))

# Calculate normalized counts
season_counts <- airline_clean %>%
  group_by(Season) %>%
  summarize(
    total_flights = n(),
    days_in_season = n_distinct(Flight_Date)
  ) %>%
  mutate(flights_per_day = total_flights / days_in_season)

# Chart 1: Raw counts (misleading)
plot_raw <- ggplot(airline_clean, aes(x = Season)) +
  geom_bar(fill = "tomato") +
  labs(
    title = "Before: Raw Flight Counts",
    subtitle = "Looks like Peak has way more flights",
    x = "Season",
    y = "Number of Flights"
  )

# Chart 2: Normalized (fair comparison)
plot_normalized <- ggplot(season_counts, aes(x = Season, y = flights_per_day)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "After: Flights per Day",
    subtitle = "Actually roughly equal across seasons",
    x = "Season",
    y = "Flights per Day"
  )

# Put them side by side
plot_raw + plot_normalized
```

The raw count makes Peak look dominant, but once normalized by days in each season, flights are actually spread pretty evenly. This is an important data integrity check before drawing conclusions from season data.

Peak (June to August, late December): Busiest travel period with the highest demand. Covers summer vacation and winter holidays. Flights are more expensive and airports are crowded.

Shoulder (April to May, September to October): Transition period between peak and low seasons. Demand and prices are moderate, with fewer crowds than peak times.

Normal (March, November): Average travel demand. Not busy or quiet, with standard pricing.

Low (January to February): Slowest travel period with the lowest demand. Flights are usually cheaper and less crowded.

## Research Questions

### 1. How does profit spread across all flights?

```{r}
ggplot(airline_clean, aes(x = Profit)) +
  geom_histogram(binwidth = 50000, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "K")) +
  labs(
    title = "Distribution of Profit Across Flights",
    x = "Profit (USD)",
    y = "Number of Flights"
  )
```

Most flights earn between $0 and $500K. The distribution is right-skewed, meaning a small number of high-performing flights earn significantly more than the rest, with the bulk of the data reaching close to $700K. There are also flights with negative profit, meaning they are operating at a loss.

### 2. Are long-haul routes more profitable than short-haul routes?

```{r}
airline_clean %>%
  group_by(Route_Category) %>%
  summarize(
    avg_profit = mean(Profit, na.rm = TRUE),
    avg_margin = mean(Profit_Margin, na.rm = TRUE)
  )
```

```{r}
# Profit vs Profit Margin by Route Category (side by side)
plot_profit_route <- ggplot(airline_clean, aes(x = Route_Category, y = Profit)) +
  geom_boxplot(fill = "tomato") +
  labs(title = "Profit by Route Category", x = "Route Category", y = "Profit (USD)")

plot_margin_route <- ggplot(airline_clean, aes(x = Route_Category, y = Profit_Margin)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Profit Margin by Route Category", x = "Route Category", y = "Profit Margin (%)")

plot_profit_route + plot_margin_route
```

Long-haul routes earn more absolute profit, but their margin is not dramatically higher. Short-haul routes are smaller in dollar terms but can be just as efficient. So yes, long-haul is more profitable in total dollars, but not because it is more efficient. Mostly it is because the revenue is larger to begin with.

### 3. Do flights generate more profit during Peak season?

```{r}
ggplot(airline_clean, aes(x = Season, y = Profit)) +
  geom_boxplot(fill = "steelblue", color = "navy") +
  labs(
    title = "Profit by Season",
    x = "Season",
    y = "Profit (USD)"
  )
```

Peak season does show higher median profit, but the boxes overlap quite a bit. This suggests that season alone does not determine whether a flight is profitable, since other factors like demand level and load factor also play a role.

### 4. How does demand level influence profitability?

```{r}
airline_clean %>%
  group_by(Demand_Level) %>%
  summarize(
    flight_count = n(),
    avg_profit = mean(Profit, na.rm = TRUE),
    median_profit = median(Profit, na.rm = TRUE),
    avg_margin = mean(Profit_Margin, na.rm = TRUE),
    median_margin = median(Profit_Margin, na.rm = TRUE)
  )

plot_profit_demand <- ggplot(airline_clean, aes(x = Demand_Level, y = Profit)) +
  geom_boxplot(fill = "tomato") +
  labs(title = "Profit by Demand Level", x = "Demand Level", y = "Profit (USD)")

plot_margin_demand <- ggplot(airline_clean, aes(x = Demand_Level, y = Profit_Margin)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Profit Margin by Demand Level", x = "Demand Level", y = "Profit Margin (%)")

plot_profit_demand + plot_margin_demand
```

This is one of the strongest findings. High demand flights are clearly more profitable in both absolute and margin terms. The separation between groups is much cleaner here than in the season analysis.

### 5. Do fuller flights lead to higher profit?

```{r}
#Load Factor VS Profit
ggplot(airline_clean, aes(x = Load_Factor, y = Profit)) +
  geom_point(alpha = 1/3, color = "steelblue") +
  labs(
    title = "Load Factor vs Profit",
    x = "Load Factor (Passengers / Aircraft Capacity)",
    y = "Profit (USD)"
  )
```

```{r}
ggplot(airline_clean, aes(x = Load_Factor, y = Profit)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Load Factor vs Profit (with Break-Even Line)",
    subtitle = "Red dashed line = break-even (Profit = $0)",
    x = "Load Factor (Passengers / Aircraft Capacity)",
    y = "Profit (USD)"
  )
```

All flights in this dataset already have a high load factor, so the data looks compact and clustered. Even then, a lot of flights still fall below the break-even line. This tells us that load factor alone does not guarantee profitability. Other fixed costs like fuel, crew salaries, and airport fees also play a significant role in determining whether a flight makes money.

### 6. Is profit more influenced by season or by demand level?

```{r}
# by season
airline_clean %>%
  group_by(Season) %>%
  summarize(
    avg_profit = mean(Profit, na.rm = TRUE),
    avg_margin = mean(Profit_Margin, na.rm = TRUE)
  )

# by demand
airline_clean %>%
  group_by(Demand_Level) %>%
  summarize(
    avg_profit = mean(Profit, na.rm = TRUE),
    avg_margin = mean(Profit_Margin, na.rm = TRUE)
  )
```

```{r}
# Profit vs Profit Margin by Season (side by side)
plot_profit_season <- ggplot(airline_clean, aes(x = Season, y = Profit)) +
  geom_boxplot(fill = "tomato") +
  labs(title = "Profit by Season", x = "Season", y = "Profit (USD)")

plot_margin_season <- ggplot(airline_clean, aes(x = Season, y = Profit_Margin)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Profit Margin by Season", x = "Season", y = "Profit Margin (%)")

plot_profit_season + plot_margin_season
```

Demand level shows a cleaner separation between groups than season does. Season has an effect, but a high demand flight in Low season can still outperform a low demand flight in Peak season. Demand level is the stronger driver.

### 7. Which season and route combinations are most profitable?

```{r}
airline_clean %>%
  group_by(Season, Route_Category) %>%
  summarize(
    avg_profit = mean(Profit, na.rm = TRUE),
    avg_margin = mean(Profit_Margin, na.rm = TRUE),
    .groups = "drop"
  )
```

Peak season combined with Long Haul consistently produces the highest average profit and margin. Low season combined with Short Haul is the weakest combination. Airlines should prioritize long-haul scheduling during peak periods.

### 8. How does profit change month by month?

```{r}
airline_clean %>%
  mutate(Month = floor_date(Flight_Date, "month")) %>%
  group_by(Month) %>%
  summarize(avg_profit = mean(Profit, na.rm = TRUE)) %>%
  ggplot(aes(x = Month, y = avg_profit)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "navy", size = 2) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Average Monthly Profit Over Time",
    subtitle = "Red dashed line = break-even",
    x = "Month",
    y = "Average Profit (USD)"
  )
```

Interestingly, average profit is actually highest in January and December, and drops to its lowest point in the summer months of June and July. This seems counterintuitive given the season labels, but it makes sense when you think about it. In peak summer months, airlines run far more total flights, including many lower-demand routes that bring the overall average down. In quieter months, only the most reliable and profitable routes tend to operate, which pushes the average up. The red dashed line confirms that no month averages below zero, meaning flights are profitable across the whole year on average.

## Conclusion

This analysis looked at what drives profitability across airline routes using 7,974 flights. A few clear patterns came out:

**Demand level is the strongest predictor of profitability.** High demand flights consistently outperform low demand ones in both profit and profit margin, more so than season alone.

**Long-haul routes earn more money, but not because they are more efficient.** When you look at profit margin instead of raw profit, the gap between route categories shrinks significantly.

**Load factor matters a lot.** Keeping seats filled is essential for covering fixed costs like fuel, crew, and airport fees. Flights with low load factors are far more likely to operate at a loss.

**The monthly trend is counterintuitive.** Average profit per flight is actually highest in winter months and lowest in summer, likely because peak season floods the schedule with lower-demand routes that drag the average down.

Overall, the data suggests that airlines should focus on maximizing demand and load factor rather than just adding long-haul routes. Efficiency matters as much as scale.
