---
title: "Airline Route Profitability"
subtitle: "DATA 2401"
author: "Niyog Uprety & Shaheer Ahmed"
date: "April 2026"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data, include=FALSE}
library(tidyverse)
library(lubridate)
library(scales)

airline_data <- read_csv("airline_route_profitability.csv")

airline_clean <- airline_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  mutate(across(where(is.character), ~ ifelse(is.na(.), "Unknown", .))) %>%
  mutate(Flight_Date = ymd(Flight_Date))
```

## About the Dataset

- **Source:** Kaggle, Airline Route Profitability Dataset by Muhammad Waleed
- **Context:** Simulated full-year financial data for a major Middle Eastern airline based in **Dubai (DXB)**
- **Size:** 7,974 flights across multiple routes, seasons, and aircraft types
- **Goal:** Find what actually drives airline profitability
- **Key metrics:** We use both **Profit** and **Profit Margin**, since raw profit alone can be misleading

---

## Route Network and Fleet

- **Short-Haul (1 to 3.5 hrs):** DOH, MCT, KWI, BAH, RUH, JED, CAI, AMM
- **Medium-Haul (2.5 to 7 hrs):** BOM, DEL, IST, LHR, BLR, HYD, MAA, KHI, LHE, CMB
- **Long-Haul (6.5 to 16.5 hrs):** JFK, SYD, MEL, SIN, HKG, BKK, KUL, CDG, FRA, LAX, SFO, ORD
- **Aircraft types:** Boeing 737-800, Airbus A320, Boeing 777-300ER, Airbus A380, Boeing 787-9, Airbus A350-900

A Dubai-based carrier operates both regional and intercontinental flights, so profitability depends on both **route scale** and **operating efficiency**.

---

## Data Cleaning

- Checked for missing values across all 33 columns
- Replaced missing numeric values with the **median**
- Replaced missing text values with **"Unknown"**
- Converted Flight_Date into proper date format
- Dataset was fairly clean with no major issues found

```{r}
cat("Missing values after cleaning:", sum(is.na(airline_clean)))
```

---

## Key Variables in the Analysis

- **Profit:** Total earnings after costs
- **Profit_Margin:** Profit as a percentage of revenue
- **Load_Factor:** How full the flight was
- **Route_Category:** Short-haul, medium-haul, or long-haul
- **Season:** Peak, Shoulder, Normal, or Low
- **Demand_Level:** High or Medium demand
- **Flight_Date:** Used for monthly trend analysis

These variables help compare both **how much money flights make** and **how efficiently they make it**.

---

## Correct Season Definitions

- **Peak:** January to February and December
- **Shoulder:** March to April and October to November
- **Normal:** May and September
- **Low:** June to August

This is important because it means **winter months are Peak** and **summer months are Low**, which explains the monthly profit trend later.

---

## How Does Profit Spread Across Flights?

```{r fig.width=9, fig.height=4.5}
ggplot(airline_clean, aes(x = Profit)) +
  geom_histogram(binwidth = 50000, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(
    title = "Distribution of Profit Across Flights",
    x = "Profit (USD)",
    y = "Number of Flights"
  ) +
  theme_minimal()
```

Most flights earn between **$0 and $500K**. Some operate at a loss, while a smaller number earn much more than the rest.

---

## Are Long-Haul Routes More Profitable?

```{r fig.width=9, fig.height=4.5}
ggplot(airline_clean, aes(x = Route_Category, y = Profit)) +
  geom_boxplot(fill = "tomato") +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(title = "Profit by Route Category", x = "Route Category", y = "Profit (USD)") +
  theme_minimal()
```

Long-haul routes generally earn the **highest total profit**, while short-haul routes perform the weakest.

---

## Are Long-Haul Routes More Efficient?

```{r fig.width=9, fig.height=4.5}
ggplot(airline_clean, aes(x = Route_Category, y = Profit_Margin)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Profit Margin by Route Category", x = "Route Category", y = "Profit Margin (%)") +
  theme_minimal()
```

Once we divide by revenue, the gap becomes smaller. Long-haul earns **more dollars**, but is **not always proportionally more efficient**.

---

## Does Season Affect Profit?

```{r fig.width=9, fig.height=4.5}
ggplot(airline_clean, aes(x = Season, y = Profit)) +
  geom_boxplot(fill = "tomato") +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(title = "Profit by Season", x = "Season", y = "Profit (USD)") +
  theme_minimal()
```

The pattern is clear: **Peak** performs best, followed by **Shoulder**, then **Normal**, with **Low** performing worst.

---

## Demand Level is the Strongest Driver

```{r fig.width=9, fig.height=4.5}
ggplot(airline_clean, aes(x = Demand_Level, y = Profit)) +
  geom_boxplot(fill = "tomato") +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(title = "Profit by Demand Level", x = "Demand Level", y = "Profit (USD)") +
  theme_minimal()
```

This is the clearest pattern in the dataset. High-demand flights are much more profitable than medium-demand flights.

---

## Demand Level Also Improves Efficiency

```{r fig.width=9, fig.height=4.5}
ggplot(airline_clean, aes(x = Demand_Level, y = Profit_Margin)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Profit Margin by Demand Level", x = "Demand Level", y = "Profit Margin (%)") +
  theme_minimal()
```

Demand level gives an even cleaner separation than season, which is why it is the strongest overall driver.

---

## Fuller Flights = More Profit

```{r fig.width=9, fig.height=4.5}
ggplot(airline_clean, aes(x = Load_Factor, y = Profit)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(
    title = "Load Factor vs Profit",
    subtitle = "Red dashed line = break-even point",
    x = "Load Factor (Passengers / Capacity)",
    y = "Profit (USD)"
  ) +
  theme_minimal()
```

All flights already have a high load factor, so the data looks compact and clustered. Even then, a lot of flights still fall below break-even. So fuller flights help, but **costs still matter**.

---

## Profit Over Time

```{r fig.width=9, fig.height=4.5}
airline_clean %>%
  mutate(Month = floor_date(Flight_Date, "month")) %>%
  group_by(Month) %>%
  summarize(avg_profit = mean(Profit, na.rm = TRUE)) %>%
  ggplot(aes(x = Month, y = avg_profit)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "navy", size = 2) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(
    title = "Average Monthly Profit Over Time",
    subtitle = "Red dashed line = break-even",
    x = "Month",
    y = "Average Profit (USD)"
  ) +
  theme_minimal()
```

Winter months perform best because they fall into **Peak season**. Summer months perform worst because they fall into **Low season**.

---

## Best Season and Route Combination

```{r}
library(knitr)
airline_clean %>%
  group_by(Season, Route_Category) %>%
  summarize(
    Avg_Profit = mean(Profit, na.rm = TRUE),
    Avg_Margin = mean(Profit_Margin, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(Avg_Profit)) %>%
  kable(digits = 2, caption = "Profitability by Season and Route Category")
```

The strongest combination is **Peak + Long Haul**, while **Low + Short Haul** is the weakest.

---

## Key Takeaways

- **Demand level** is the strongest predictor of profitability
- **Long-haul routes** earn the most total profit, but are not always the most efficient
- **Season clearly matters** once the correct definitions are used. Peak = winter, Low = summer
- **Load factor helps**, but it does not guarantee profit on its own
- Best overall pattern: **Peak season + Long Haul**

---

## Limitations and Future Work

**Limitations:**

- This dataset is **simulated**, not taken directly from a real airline
- It reflects one airline setting based in **Dubai**
- Results may not generalize to all airlines or regions

**Future work could explore:**

- Profitability by **aircraft type**
- Profitability by **specific destination**
- Predictive modeling for flight profit
