---
title: "Airline Route Profitability"
output: html_document
date: "2026-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
The dataset used in this project is the "Airline Route Profitability Dataset" 
from Kaggle, available at: https://www.kaggle.com/datasets/waleedfaheem/airline-route-profitability-and-cost-analysis?resource=download
Author: Muhammad Waleed 
```{r}
library(tidyverse)
library(tidyr)
library(lubridate)
library(patchwork)
```

```{r}
#Loading Data
airline_data <- read_csv("airline_route_profitability.csv")
glimpse(airline_data)
```

```{r}
head(airline_data)
```

```{r}
# Check missing values before
colSums(is.na(airline_data))

# Replace NAs in numeric columns with the median
airline_clean <- airline_data %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Replace NAs in text columns with "Unknown"
airline_clean <- airline_clean %>%
  mutate(across(where(is.character),
                ~ ifelse(is.na(.), "Unknown", .)))

# Check missing values after
colSums(is.na(airline_clean))

```
```{r}
glimpse(airline_clean)
```

```{r}
#Average profit by season
airline_clean %>%
  group_by(Season) %>%
  summarize(avg_profit = mean(Profit, na.rm = TRUE))
#Peak = busiest (summer/holidays)
#Shoulder = in between peak and low
#Normal = regular
#Low = slowest
```

```{r}
# Average profit by route category
airline_clean %>%
  group_by(Route_Category) %>%
  summarize(avg_profit = mean(Profit, na.rm = TRUE))

#Short Haul = 1 to 3.5 hours
#Medium Haul = 2.5 to 7 hours
#Long Haul = 6.5 to 16.5 hours
```


```{r}
#Average profit by demand level
airline_clean %>%
  group_by(Demand_Level) %>%
  summarize(avg_profit = mean(Profit, na.rm = TRUE))
```

```{r}
# Convert Flight_Date to a proper date
airline_clean <- airline_clean %>%
  mutate(Flight_Date = ymd(Flight_Date))

# Calculate normalized counts
season_counts <- airline_clean %>%
  group_by(Season) %>%
  summarize(
    total_flights = n(),
    days_in_season = n_distinct(Flight_Date)
  ) %>%
  mutate(flights_per_day = total_flights / days_in_season)

# Chart 1: Raw counts (misleading)
plot_raw <- ggplot(airline_clean, aes(x = Season)) +
  geom_bar(fill = "tomato") +
  labs(
    title = "Before: Raw Flight Counts",
    subtitle = "Looks like Peak has way more flights",
    x = "Season",
    y = "Number of Flights"
  )

# Chart 2: Normalized (fair comparison)
plot_normalized <- ggplot(season_counts, aes(x = Season, y = flights_per_day)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "After: Flights per Day",
    subtitle = "Actually roughly equal across seasons",
    x = "Season",
    y = "Flights per Day"
  )

# Put them side by side
plot_raw + plot_normalized
```
Peak (June–August, late December): Busiest travel period with the highest demand. Covers summer vacation and winter holidays. Flights are more expensive and airports are crowded.

Shoulder (April–May, September–October): Transition period between peak and low seasons. Demand and prices are moderate, with fewer crowds than peak times.

Normal (March, November): Average travel demand. Not busy or quiet, with standard pricing.

Low (January–February): Slowest travel period with the lowest demand. Flights are usually cheaper and less crowded.

```{r}
#How profit spreads out across flights?

ggplot(airline_clean, aes(x = Profit)) +
  geom_histogram(binwidth = 50000) +
  labs(
    title = "Distribution of Profit",
    x = "Profit (USD)",
    y = "Number of Flights"
  )
```
x-axis values are in dollars. 0e+00 = $0, 5e+05 = $500,000, 1e+06 = $1,000,000

```{r}
#Profit by Route Category
ggplot(airline_clean, aes(x = Route_Category, y = Profit)) +
  geom_boxplot() +
  labs(
    title = "Profit by Route Category",
    x = "Route Category",
    y = "Profit (USD)"
  )
```

```{r}
#Load Factor VS Profit
ggplot(airline_clean, aes(x = Load_Factor, y = Profit)) +
  geom_point(alpha = 1/3) +
  labs(
    title = "Load Factor vs Profit",
    x = "Load Factor",
    y = "Profit (USD)"
  )
#Load_Factor = Passengers/Aircraft Capacity
```

```{r}
#Do flights generate more profit during Peak season compared to Low or Shoulder seasons?
ggplot(airline_clean, aes(x = Season, y = Profit)) +
  geom_boxplot() +
  labs(title = "Profit by Season")
```

```{r}
#Are long-haul routes more profitable than short-haul routes?
airline_clean %>%
  group_by(Route_Category) %>%
  summarize(avg_profit = mean(Profit))
```

```{r}
#How does demand level influence airline profitability?
# Average profit by demand level
airline_clean %>%
  group_by(Demand_Level) %>%
  summarize(
    flight_count = n(),
    avg_profit = mean(Profit, na.rm = TRUE),
    median_profit = median(Profit, na.rm = TRUE),
    min_profit = min(Profit, na.rm = TRUE),
    max_profit = max(Profit, na.rm = TRUE)
  )

ggplot(airline_clean, aes(x = Demand_Level, y = Profit)) +
  geom_boxplot() +
  labs(
    title = "Profit by Demand Level",
    x = "Demand Level",
    y = "Profit (USD)"
  )
```

```{r}
#Do fuller flights (higher load factor) lead to higher profit?
ggplot(airline_clean, aes(x = Load_Factor, y = Profit)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm")
```

```{r}
#Is profit more influenced by season or by demand level?
# by season
airline_clean %>%
  group_by(Season) %>%
  summarize(avg_profit = mean(Profit))

# by demand
airline_clean %>%
  group_by(Demand_Level) %>%
  summarize(avg_profit = mean(Profit))
```

```{r}
#Which combinations of season and route category produce the highest profit?
airline_clean %>%
  group_by(Season, Route_Category) %>%
  summarize(avg_profit = mean(Profit))
```
